### 2022-06-23 manuscript/hiv-gwas.tex
Update on Overleaf.

\begin{flushright} % for MBE submission
Submission type: Article\\
Section: Methods
\end{flushright}

@@

% Figure 1 location

@@

% Figure 2-4 location

@@

% Figure 5 location

@@

% Figure 6 location

% Table 1 location

@@

% Figure 7 location

@@

% Main text table

% Main text figure legends

\begin{figure}[H]
	\begin{center}
% 	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
	\caption{A high-level schematic of our phylogenetic Ornstein-Uhlenbeck mixed model (POUMM)-based simulation framework in the context of HIV-1 set-point viral load (spVL). (A) shows how the viral effects on spVL evolve along the viral phylogeny according to an Ornstein-Uhlenbeck process. (B) shows how human host genetic effects are the sum of independent effects from several causal variants. Each variant can be present in 0, 1, or 2 copies. Half the variants have a positive effect of size $\delta$ and half have a negative effect of size $\delta$. (C) shows how other environmental effects are independently drawn from a Gaussian distribution centered at 0. These three effects sum to the trait value for each simulated individual.}
		\label{fig:spVLModel}
	\end{center}
\end{figure}

\begin{figure}[H]
	\begin{center}
% 		\includegraphics[width=\linewidth]{figures/simulation_results}
		\caption{Results from the simulation study. We simulated host, pathogen, and environmental effects on a trait under the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) with different heritability ($H^2$; y-axis) and selection strength ($\alpha$; x-axis) parameters. For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values. (A) shows the root mean squared error (RMSE) of our estimator (left) compared to un-corrected trait values, scaled by their mean (right) under each simulated evolutionary scenario. The RMSE is with reference to the true (simulated) host part of the trait values. Thus, more accurate estimates (lower RMSE) mean the trait value used for GWAS will be closer to the true host part of the trait value. (B) shows how genome-wide association study (GWAS) power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the scaled trait value (right). Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples. The points highlight specific heritability and selection strength values from the \emph{A. thaliana}-\emph{X. arboricola} quantitative disease resistance (QDR) analysis, HIV-1 spVL analysis, and four simulated scenarios that are presented in more detail in Figure \ref{fig:tip-trait-correlation}.}
		\label{fig:simulationResults}
	\end{center}
\end{figure}

\begin{figure}[H]
	\begin{center}
% 		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values.pdf}
		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves genome-wide association study (GWAS) power (right side) and where it does not (left side). These examples correspond to two of the unfilled points in Figure \ref{fig:simulationResults}. (A) and (B) show total trait values for 12 randomly selected tips from the simulated phylogeny with pathogen heritability $H^2$ of 15 and 75\%, respectively. Depending on the pathogen heritability, trait values are more or less correlated at clustered tips. (C) compares our method's estimate for the non-pathogen part of trait values (y-axis) with true simulated host trait values (x-axis) with pathogen heritability of 15 and 75\%. The solid line is the y=x line. Selection strength $\alpha$ was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
		\label{fig:simulationTraitExample}
	\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
    % \includegraphics[width = \linewidth]{figures/tip_trait_correlation.pdf}
	\caption{Correlations between trait values in pairs of tips in four simulated scenarios. These examples correspond to the four unfilled points in Figure \ref{fig:simulationResults}. Correlations are calculated for pairs of tips binned by phylogenetic distance (into deciles) across the 20 replicate simulations for each of the four evolutionary scenarios. Trait values are only noticeably correlated for closely clustered tips under the scenario with high pathogen heritability $H^2$ and low selection strength $\alpha$/ low stochastic fluctuations $\sigma$ (upper left facet).}
	\label{fig:tip-trait-correlation}
	\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
    % \includegraphics[width = \linewidth]{figures/application_experimental_setups.pdf}
	\caption{A high-level schematic of the experimental setup for the two application datasets. For (A) HIV-1 set-point viral load (spVL) in the Swiss HIV Cohort Study, data are paired viral and human genotypes and associated spVL measurements. We fit the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) to the viral phylogeny and spVL values associated with each infected individual ($z_1$, $z_2$, ..., $z_{1493}$). For (B) \emph{A. thaliana}-\emph{X. arboricola} quantitative disease resistance (QDR) from \cite{Wang2018Two-wayGenomes}, data are bacterial and plant genotypes with QDR measurements for all possible combinations of pathogen and host plant strains. We fit the POUMM to the bacterial phylogeny and mean QDR calculated for each pathogen strain across all the hosts plant types ($\bar{z}_1$, $\bar{z}_2$, ..., $\bar{z}_{22}$).}
	\label{fig:application-exp-setup}
	\end{center}
\end{figure}

\begin{figure}[H]
	\begin{center}
% 		\includegraphics[width=\linewidth]{figures/gwas_results}
		\caption{Results from comparative genome-wide association studies (GWAS) on HIV-1 set-point viral load (spVL) data. (A) shows association p-values for the same host variants from the Swiss HIV cohort in GWAS with two different response variables. On the left, we used unmodified (total) spVL values. On the right, we used our estimates for the non-pathogen effects on spVL. The alternating shades correspond to different chromosomes. (B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively). Base positions are with reference to genome build GRCh37. The color of each point represents the difference in -log$_{10}$ p-value between the two GWAS. Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it. The dashed lines show genome-wide significance at p = $5 \times 10^{-8}$.}
		\label{fig:gwas-results}
	\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
    % \includegraphics[width = \linewidth]{figures/gwas_results_xanthomonas_maxmaf.pdf}
	\caption{Results from comparative genome-wide association studies (GWAS) on \emph{A. thaliana} quantitative disease resistance (QDR) to \emph{X. arboricola}. The two facets show association p-values for the same host \emph{A. thaliana} variants in GWAS with two different response variables. On the left, we used unmodified (total) QDR values for each of the 22 selected host-pathogen pairings on which these results are based. On the right, we used our estimates for the non-pathogen effects on QDR for these samples. In this case, estimated non-pathogen effects are the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A. thaliana} types. The alternating shades correspond to different chromosomes. The dashed lines show significance at significance level 0.05 with a Bonferroni correction for multiple testing.}
	\label{fig:gwas-results-qdr}
	\end{center}
\end{figure}

>>>>

% \begin{flushright} % for MBE submission
% Submission type: Article\\
% Section: Methods
% \end{flushright}

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
	\caption{A high-level schematic of our phylogenetic Ornstein-Uhlenbeck mixed model (POUMM)-based simulation framework in the context of HIV-1 set-point viral load (spVL). (A) shows how the viral effects on spVL evolve along the viral phylogeny according to an Ornstein-Uhlenbeck process. (B) shows how human host genetic effects are the sum of independent effects from several causal variants. Each variant can be present in 0, 1, or 2 copies. Half the variants have a positive effect of size $\delta$ and half have a negative effect of size $\delta$. (C) shows how other environmental effects are independently drawn from a Gaussian distribution centered at 0. These three effects sum to the trait value for each simulated individual.}
		\label{fig:spVLModel}
	\end{center}
\end{figure}

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\linewidth]{figures/simulation_results}
		\caption{Results from the simulation study. We simulated host, pathogen, and environmental effects on a trait under the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) with different heritability ($H^2$; y-axis) and selection strength ($\alpha$; x-axis) parameters. For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values. (A) shows the root mean squared error (RMSE) of our estimator (left) compared to un-corrected trait values, scaled by their mean (right) under each simulated evolutionary scenario. The RMSE is with reference to the true (simulated) host part of the trait values. Thus, more accurate estimates (lower RMSE) mean the trait value used for GWAS will be closer to the true host part of the trait value. (B) shows how genome-wide association study (GWAS) power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the scaled trait value (right). Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples. The points highlight specific heritability and selection strength values from the \emph{A. thaliana}-\emph{X. arboricola} quantitative disease resistance (QDR) analysis, HIV-1 spVL analysis, and four simulated scenarios that are presented in more detail in Figure \ref{fig:tip-trait-correlation}.}
		\label{fig:simulationResults}
	\end{center}
\end{figure}

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values.pdf}
		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves genome-wide association study (GWAS) power (right side) and where it does not (left side). These examples correspond to two of the unfilled points in Figure \ref{fig:simulationResults}. (A) and (B) show total trait values for 12 randomly selected tips from the simulated phylogeny with pathogen heritability $H^2$ of 15 and 75\%, respectively. Depending on the pathogen heritability, trait values are more or less correlated at clustered tips. (C) compares our method's estimate for the non-pathogen part of trait values (y-axis) with true simulated host trait values (x-axis) with pathogen heritability of 15 and 75\%. The solid line is the y=x line. Selection strength $\alpha$ was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
		\label{fig:simulationTraitExample}
	\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
    \includegraphics[width = \linewidth]{figures/tip_trait_correlation.pdf}
	\caption{Correlations between trait values in pairs of tips in four simulated scenarios. These examples correspond to the four unfilled points in Figure \ref{fig:simulationResults}. Correlations are calculated for pairs of tips binned by phylogenetic distance (into deciles) across the 20 replicate simulations for each of the four evolutionary scenarios. Trait values are only noticeably correlated for closely clustered tips under the scenario with high pathogen heritability $H^2$ and low selection strength $\alpha$/ low stochastic fluctuations $\sigma$ (upper left facet).}
	\label{fig:tip-trait-correlation}
	\end{center}
\end{figure}

@@

\begin{figure}[H]
\begin{center}
    \includegraphics[width = \linewidth]{figures/application_experimental_setups.pdf}
	\caption{A high-level schematic of the experimental setup for the two application datasets. For (A) HIV-1 set-point viral load (spVL) in the Swiss HIV Cohort Study, data are paired viral and human genotypes and associated spVL measurements. We fit the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) to the viral phylogeny and spVL values associated with each infected individual ($z_1$, $z_2$, ..., $z_{1493}$). For (B) \emph{A. thaliana}-\emph{X. arboricola} quantitative disease resistance (QDR) from \cite{Wang2018Two-wayGenomes}, data are bacterial and plant genotypes with QDR measurements for all possible combinations of pathogen and host plant strains. We fit the POUMM to the bacterial phylogeny and mean QDR calculated for each pathogen strain across all the hosts plant types ($\bar{z}_1$, $\bar{z}_2$, ..., $\bar{z}_{22}$).}
	\label{fig:application-exp-setup}
	\end{center}
\end{figure}

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\linewidth]{figures/gwas_results}
		\caption{Results from comparative genome-wide association studies (GWAS) on HIV-1 set-point viral load (spVL) data. (A) shows association p-values for the same host variants from the Swiss HIV cohort in GWAS with two different response variables. On the left, we used unmodified (total) spVL values. On the right, we used our estimates for the non-pathogen effects on spVL. The alternating shades correspond to different chromosomes. (B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively). Base positions are with reference to genome build GRCh37. The color of each point represents the difference in -log$_{10}$ p-value between the two GWAS. Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it. The dashed lines show genome-wide significance at p = $5 \times 10^{-8}$.}
		\label{fig:gwas-results}
	\end{center}
\end{figure}

@@

\begin{figure}[H]
\begin{center}
    \includegraphics[width = \linewidth]{figures/gwas_results_xanthomonas_maxmaf.pdf}
	\caption{Results from comparative genome-wide association studies (GWAS) on \emph{A. thaliana} quantitative disease resistance (QDR) to \emph{X. arboricola}. The two facets show association p-values for the same host \emph{A. thaliana} variants in GWAS with two different response variables. On the left, we used unmodified (total) QDR values for each of the 22 selected host-pathogen pairings on which these results are based. On the right, we used our estimates for the non-pathogen effects on QDR for these samples. In this case, estimated non-pathogen effects are the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A. thaliana} types. The alternating shades correspond to different chromosomes. The dashed lines show significance at significance level 0.05 with a Bonferroni correction for multiple testing.}
	\label{fig:gwas-results-qdr}
	\end{center}
\end{figure}

@@
